---
title: "5301 Final Project"
author: "Hariharan Selvam"
date: "`r Sys.Date()`"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## TOPIC: TWO-WAY ANOVA


## About the dataset:
It is collected from r packages.
It contains measurements of trunk circumference (in millimeters) for five orange trees over a span of seven time points
The dataset captures the growth of these trees over time

There are three different features.
1. Tree: A factor indicating the tree number (1 to 5).
2. Age: The age of the tree in days since December 31, 1968.
3. circumference: Trunk circumference of the tree in millimeters.

## Importing the dataset:

```{r }
#Importing the dataset
data("Orange")
Orange
df <- data.frame(Orange)
df
df$Tree <- as.factor(df$Tree)
df$Tree <- factor(df$Tree)



df_numeric <- data.frame(Orange)
df_numeric$Tree <- as.numeric(as.character(df$Tree))
df_numeric$age <- as.numeric(as.character(df$age))

str(df$Tree)
str(df$age)
summary(df$Tree)
summary(df$age)
summary(df)
summary(is.na(df))
```
```{r}
head(df)
```

## Null and Alternate hypotheses:
In two-way ANOVA, there will be 3 different null and alternate hypotheses.

case 1:
Null hypothesis: Mean factor of column Tree is same
Alternate hypothesis: Mean factor of column Tree is not same

case2:
Null hypothesis: Mean factor of column age is same
Alternate hypothesis: Mean factor of column age is not same

Case3:
Null Hypothesis: There is no interaction between the factor Tree and age
Alternate Hypothesis: There is interaction between the factor Tree and age

Here we're taking the significant value as 0.05

## Descriptive Data Analysis and Exploratory Data Analysis:

```{r}
# Summary statistics for the entire dataset
summary(df)

# Summary statistics for specific variables
summary(df$age)
summary(df$Tree)
summary(df$circumference)

```

## Histogram:

Here we've plotted histogram for numerical variable 'circumference'
The following r code is used to perform.

hist(df$circumference, main = "Circumference Distribution", xlab = "Circumference")


```{r}
# Histogram for a numeric variable
hist(df$circumference, main = "Circumference Distribution", xlab = "Circumference")

```
Here we found that the histogram is,
Unimodal distribution: The peak at around 150 units of circumference indicates that this is the most common circumference value.

Approximately symmetrical distribution: The roughly equal left and right halves suggest a normal distribution.

Majority of values clustered around 150: The majority of the circumference values fall within the range centered around 150.

Mode at 150: The most common circumference value is 150.

Median near 150: Half of the values are above 150 and half below.

Range from 50 to 200: The data set spans a range of 150 units.

## Statistical parameters for the numerical variable


```{r}
# Mean and median
mean(df$circumference)
median(df$circumference)

# Standard deviation and interquartile range
sd(df$circumference)
IQR(df$circumference)

```
From the above code we got to know that,
The mean value for circumference is 115.8571.
The median value for circumference is  115.
The standard deviation value for circumference is 57.48818.
The interquartile range value for circumference is 96.

## Correlation matrix:

```{r}

cor_matrix <- cor(df_numeric[, c("age", "Tree", "circumference")])
print(cor_matrix)
library(corrplot)
corrplot(cor_matrix, method = "color")

```
From the correlation matrix:

The correlation between age and Tree is 0.8, which indicates a strong positive relationship. 

The correlation between Tree and circumference is 0.9, which also indicates a strong positive relationship. 

The correlation between age and circumference is 0.7, which indicates a moderate positive relationship. 

Overall, the correlation matrix shows that there is a strong positive relationship between all three variables.

## Boxplot:

```{r}
boxplot(circumference ~ Tree, data = df, main = "Circumference by Tree", xlab = "Tree", ylab = "Circumference")
```

From the boxplot,
The median circumference value is highest for tree 5, followed by tree 4 and tree 3.

The IQR is smallest for tree 3 and largest for tree 1. This indicates that the circumference values for tree 3 are more tightly clustered around the median, while the circumference values for tree 1 are more spread out.

There are no outliers detected in any of the groups.



## Assumptions:

## Shapiro- Wilk test for normality:


```{r }
shapiro_test <- shapiro.test(Orange$circumference)
print(shapiro_test)

qqnorm(Orange$circumference)
qqline(Orange$circumference, col="green")

```
here our p-value >0.05 , so we can conclude that the data follows normal distribution
And also here we've failed to reject the null hypothesis.


## Barlette test for Homogeneity of variance:
```{r}
barlett_test <- bartlett.test(circumference ~ Tree, data = df)
print(barlett_test)
```
Here our p-value >0.05 , so we can conclude that variances are equal across the groups.

## Two way ANOVA

```{r}
#Two-way ANOVA
two_way_anova <- aov(circumference ~ age + Tree + age:Tree, data = df)
summary(two_way_anova)
```
From this two way ANOVA test, we got to know

## Main Effect of 'age':

Degrees of Freedom (Df): 1
Sum of Squares (Sum Sq): 93771.54
Mean Square (Mean Sq): 93771.54
F-value: 864.735
p-value (Pr(>F)): < 2e-16 (extremely small)
The main effect of 'age' is highly significant, suggesting that the mean 'circumference' significantly differs across different levels of 'age'.

## Main Effect of 'Tree':
  
Degrees of Freedom (Df): 4
Sum of Squares (Sum Sq): 11840.86
Mean Square (Mean Sq): 2960.22
F-value: 27.298
p-value (Pr(>F)): 8.43e-09 (extremely small)
The main effect of 'Tree' is highly significant, indicating that the mean 'circumference' significantly differs across different levels of 'Tree'.

##Interaction Effect ('age:Tree'):
  
Degrees of Freedom (Df): 4
Sum of Squares (Sum Sq): 4042.90
Mean Square (Mean Sq): 1010.72
F-value: 9.321
p-value (Pr(>F)): 9.40e-05 (extremely small)
The interaction effect between 'age' and 'Tree' is highly significant, suggesting that the effect of 'age' on 'circumference' is not consistent across all levels of 'Tree', and vice versa.

# Residuals:
Degrees of Freedom (Df): 25
Sum of Squares (Sum Sq): 2710.99
Mean Square (Mean Sq): 108.44
The residuals represent the unexplained variation in 'circumference' that is not accounted for by 'age', 'Tree', or their interaction.

#Residual Standard Error:
Residual standard error: 10.41344
This provides an estimate of the standard deviation of the residuals.

Here post hoc analysis is required.

## Post- Hoc Analysis:

```{r}
##post-hoc analysis:
library(agricolae)
tukey_result <- TukeyHSD(two_way_anova)
print(tukey_result)

plot(tukey_result)
```
From the post- hoc analysis,

## Conclusion:

The post-hoc analysis using TukeyHSD shows that there are statistically significant differences in the mean levels of tree differences between the following pairs of treatments:
  
4 and 5
2 and 5
4 and 1
2 and 1
5 and 1

The differences in the mean levels of tree differences between the following pairs of treatments are not statistically significant:
4 and 3
2 and 3
5 and 3
1 and 3

## Conclusion:
The project provides valuable insights into the growth patterns of orange trees. Both 'age' and 'Tree' significantly  impact the trunk circumference, with varying effects across different tree groups. Post-hoc analysis pinpointed specific pairs of tree groups with significant differences, offering detailed information about the distinct growth patterns.

Here, we’ve failed to reject the null hypothesis.

Therefore,
The mean circumference is the same across all levels of the "Tree" factor.
The mean circumference is the same across all levels of the "Age" factor.
There is no interaction effect between the "Tree" and "Age" factors on the mean circumference
















